Searching for long-period transiting planets in the \kepler\ light curves using supervised classification

### 2015-04-09 document/ms.tex
WORDS IN UR FACE TEHSIS

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\KT}{\project{K2}}
\newcommand{\tess}{\project{TESS}}
\newcommand{\plato}{\project{PLATO}}
\newcommand{\jwst}{\project{JWST}}
\newcommand{\terra}{\project{TERRA}}
\newcommand{\pdc}{\project{PDC}}
\newcommand{\license}{MIT License}
\newcommand{\projectname}{\project{ketu}}

@@

The transit method of exoplanet detection and characterization has been
demonstrated as a powerful method for building systematic catalogs of
exoplanets.
Despite the great success of the \kepler\ Mission with thousands of planet
discoveries \dfmtodo{cite}, current methods for exoplanet discoveries are
currently limited in the range of orbital periods that can be studied.
Specifically, the standard transit search procedures only discover signals
with at least three observed transits \dfmtodo{cite Jenkins, Petigura, \etc}.
For \kepler, with a baseline of about four years, this sets an upper limit on
the detectable periods of just over a year.
In the Solar System, Jupiter---with a period of 12 years---dominates the
planetary dynamics and, since it would only exhibit at most one transit in the
\kepler\ data, it would be missed by most existing transit search procedures.
It is possible to discover long-period planets like this using targeted radial
velocity (RV) surveys \dfmtodo{cite} but the cost of implementing a systematic
RV search is substantially higher than searching the existing and forthcoming
photometric data for single transits.

@@

Nearly every transit search algorithm is built on the same principles and many
of the same decisions are made.
In particular, at the heart of most methods is a matched filtering step where
the likelihood of an approximate transit model is computed on a grid in the
physical parameters \dfmtodo{cite}.
Using these methods, and substantial hand-curation, some long-period
transiting candidates have been published \citep[for example][]{Batalha:2013,
Huang:2013, Kipping:2014a}.
For more recent data releases, the community has settled on more conservative
selection criteria where candidates are required to have multiple transits
\citep[for example][]{Petigura:2013, Burke:2014}.
Even the \project{QATS} algorithm \dfmtodo{cite} for finding quasiperiodic
transits builds on much the same infrastructure.
This means that there has never been a systematic search for single transits
in the full \kepler\ dataset.

@@

Automated classification processes have been hugely successful in the machine
learning literature when applied to the right datasets \dfmtodo{cite some
things}.
Many astronomical problems can't be naturally solved by the standard machine
learning models because of their heterogeneous character and heteroscedastic
uncertainties.
The search for single transits, however, is an excellent use case when applied
with care.
The noise in the light curve of a given star is relatively consistent for the
full baseline of observation and we have a very good physical generative model
for the transit signal so we can simulate arbitrarily large sets of training
examples.

@@

Assuming the stellar parameters provided by the NASA Exoplanet
Archive\footnote{\dfmtodo{URL for kepler stellar table; downloaded on SOME
DATE.}} \dfmtodo{cite Huber} and approximating the stellar noise using the
15-hour CDPP \dfmtodo{cite CDPP}, \dfmtodo{some figure} shows the extrapolated
number of transiting planets with $1500\,\mathrm{d} < \period <
5000\,\mathrm{d}$ based on the \citet{Dong:2013} power-law model.
For a detection threshold $f\sim10$, the expected number of single transit
events in the \kepler\ light curves is $\sim 60$.

This results is much more optimistic than the pre-launch estimate from
\citet{Yee:2008} for a few reasons.
One effect is that, the \kepler\ Mission ran for more than four
years---longer than the fiducial Mission goal---and $\sim190,000$ stars were
targeted for nearly the full baseline instead of the original 100,000.
\dfmtodo{Explain other reasons.}

@@

The Random Forest (RF) classification model \citep{Breiman:2001} is a popular
model in the machine learning community where it is a common go-to model for
basic classification tasks.
It has also been applied with great success in astronomy \citep[for
example]{Richards:2011, Richards:2012} \dfmtodo{others}.
The RF model works by fitting an ensemble of decision trees to randomly
selected subsamples of the training dataset.
Each tree in the forest is ``grown'' by greedily choosing decision boundaries
that optimize the separation of the training data in randomly selected
subsets of the features until the separation is complete, each leaf contains
only one class or, at most, a fixed number of samples.
At each branch, the decision boundary is set by maximizing either the entropy
or the ``gini'' coefficient.

@@

To generate the \texttt{transit} examples, we simulate the transit signal
induced by the physical orbit of a large planet.
The specific distribution of simulation parameters is listed in \dfmtodo{some
table}.
In these simulations, we take limb darkening and integration time into
account\footnote{\url{https://github.com/dfm/transit}}
\citep{Mandel:2002, Kipping:2010}.
For the purposes of this \paper, we only simulate circular orbits under the
assumption that a single transit of a planet on an elliptical orbit would be
nearly identical to a circular orbit at a different period.

@@

@@

\dfmtodo{make a figure}

@@

In many supervised classification problems, a sophisticated feature
extraction method is used to improve the performance of the procedure.
For this specific problem, we find excellent performance using the 201
normalized flux values as the features.
The depth of the transit and the noise in the light curve is not irrelevant
to detection so we don't normalize the features to unit variance in the
standard way.
Instead, we normalize each sample (feature vector) by it's empirical median
value and take the logarithm.
Another option for a feature set could be the Fourier transform or wavelet
transform of the light curve section.
Alternatively, the spectrogram of the fluxes in a larger window.
We have experimented briefly with this options finding similar performance.

@@

As discussed previously, recall is the probability, integrated across the full
parameter space of interest that a true transit signal will be detected by the
method.
In astronomy, this quantity is generally called the completeness or detection
efficiency and it is routinely measured for transit surveys \dfmtodo{cite}.
On the other hand, precision is the probability that a positive classification
will be true.
In astronomy, precision is generally hard to measure because we rarely have
access to the ground truth.
For this problem, correctly computing the recall is impossible because we
never know that \emph{there is definitely no transit} at a given time in the
light curve.

@@

{\item \emph{the range of physical parameters used to simulate the transit
signals} --- The search will be most sensitive to transits with parameters
spanned by the training set but training on too many low signal-to-noise
examples in noisy light curves leads to decreased precision.
The parameter ranges used for the demonstrations here are given in
\dfmtodo{some table}.}

@@

We estimate the precision of each RF classification model on the validation
set as described in \sect{train}.
As discussed in that \sectionname, we choose a target precision to
100~percent and set the score threshold accordingly.
This estimate is only applicable to the test set under two major assumptions.
The first is that all transiting planets with more than one transit and a
large signal-to-noise ratio have been previously discovered by one of the many
transit searches applied the \kepler\ dataset \dfmtodo{cite}.
This is a safe assumption because, for large planets, these surveys are all
largely complete out to periods yielding only 2 or 3 transits in the \kepler\
baseline \dfmtodo{cite}.
The second assumption is not always valid so it makes this estimate of the
precision only approximate and not completely reliable.
This assumption is that the noise processes in the data are stationary.
In other words, the transit-like noise signals in the training and validation
sets must be ``similar'' to the signals in the test section.
This constraint is nearly satisfied when the PDC light curves are used
(\sect{data}) and the data are split carefully (\sect{split}).
As demonstrated by the mis-classifications shown in \sect{demo}, however, it
is clear that violations to this rule do exist in the data and final vetting
of the candidates must be applied using a physical understanding of the
signals.

@@

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/candidates.pdf}
\end{center}
\caption{%
\figlabel{candidates}}
\end{figure}

@@

In the 3500 light curves we searched for this \paper, we discovered one
convincing single transit signal at $830.8093\pm0.0002$~KBJD in the light
curve of the 14.9~Kep-mag G-dwarf KIC 10602068.
Despite the fact that this is a very large signal, its discovery has never
been reported.
Given the estimate of $\sim 60$ detectable single transits in the full
\kepler\ archival dataset, we expect about one discovery in 3500 light curves.

Even though only one transit is observed, it is still possible to place
constraints on the physical properties---even the orbital period---of this
transiting body.
This technique is similar in spirit to the ``photoeccentric effect''
\citep{Dawson:2012} and ``asterodensity profiling'' \citep{Kipping:2012}.
The fundamental principle is that the transit duration is a measurement of
the instantaneous velocity of the planetary motion and this yields a
constraint on the orbital period when combining this with a measurement of the
stellar density and a prior on the eccentricity of the orbit.

The \citet{Huber:2014} parameters for this star place it as a G-dwarf with a
radius of $\sim 0.9\,R_\odot$
$2.1 \pm 0.4\,R_\mathrm{J}$.

In practice, we directly model the light curve as being generated by a planet
on an eccentric Keplerian orbit transiting a limb darkened star and
marginalize over all the orbital elements and physical parameters.
We take limb darkening and the finite exposure time into account
\citep{Mandel:2002, Kipping:2010, Kipping:2013} and use an empirical
Beta function prior on the eccentricity \citep{Kipping:2013a}.

@@

>>>>

\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\KT}{\project{K2}}
\newcommand{\tess}{\project{TESS}}
\newcommand{\plato}{\project{PLATO}}
\newcommand{\jwst}{\project{JWST}}
\newcommand{\terra}{\project{TERRA}}
\newcommand{\pdc}{\project{PDC}}
\newcommand{\license}{MIT License}

@@

The transit method of exoplanet detection and characterization has been
demonstrated as a powerful method for building systematic catalogs of
exoplanets.
Despite the great success of the \kepler\ Mission with thousands of planet
discoveries \citep{Burke:2014}, current methods for exoplanet discoveries are
currently limited in the range of orbital periods that can be studied.
Specifically, the standard transit search procedures only discover signals
with at least three observed transits \citep[for example][]{Petigura:2013,
Burke:2014}.
For \kepler, with a baseline of about four years, this sets an upper limit on
the detectable periods of just over a year.
In the Solar System, Jupiter---with a period of 12 years---dominates the
planetary dynamics and, since it would only exhibit at most one transit in the
\kepler\ data, it would be missed by most existing transit search procedures.
It is possible to discover long-period planets like this using targeted radial
velocity (RV) surveys \citep[for example][]{Butler:2006, Knutson:2014} but the
cost of implementing a systematic RV search is substantially higher than
searching the existing and forthcoming photometric data for single transits.

@@

Nearly every transit search algorithm is built on the same principles and many
of the same decisions are made.
In particular, at the heart of most methods is a matched filtering step where
the likelihood of an approximate transit model is computed on a grid in the
physical parameters \citep[\kepler\ Data Processing
Handbook\footnote{\url{https://archive.stsci.edu/kepler/manuals/KSCI-19081-001_Data_Processing_Handbook.pdf}};][]{%
Petigura:2013, Huang:2013, Dressing:2015, Foreman-Mackey:2015}.
Using these methods, and substantial hand-curation, some long-period
transiting candidates have been published \citep[for example][]{Batalha:2013,
Huang:2013, Kipping:2014a}.
For more recent data releases, the community has settled on more conservative
selection criteria where candidates are required to have multiple transits
\citep[for example][]{Petigura:2013, Burke:2014}.
Even the \project{QATS} algorithm \citep{Carter:2013} for finding
quasiperiodic transits builds on much the same infrastructure.
This means that there has never been a systematic search for single transits
in the full \kepler\ dataset.

@@

Automated classification processes have been hugely successful in the machine
learning literature.
% when applied to the right datasets \dfmtodo{cite some things}.
Many astronomical problems can't be naturally solved by the standard machine
learning models because of their heterogeneous character and heteroscedastic
uncertainties.
The search for single transits, however, is an excellent use case when applied
with care.
The noise in the light curve of a given star is relatively consistent for the
full baseline of observation and we have a very good physical generative model
for the transit signal so we can simulate arbitrarily large sets of training
examples.

@@

Assuming the stellar parameters provided by the NASA Exoplanet
Archive\footnote{We downloaded the \texttt{q1\_q16\_stellar} table from
\url{http://exoplanetarchive.ipac.caltech.edu/} on 2015-04-03.}
\citep{Huber:2014} and approximating the stellar noise using the 15-hour CDPP
\citep{Christiansen:2012}, \fig{predict} shows the extrapolated number of
transiting planets with $1500\,\mathrm{d} < \period < 5000\,\mathrm{d}$ based
on the \citet{Dong:2013} power-law model.
For a detection threshold $f\sim10$, the expected number of single transit
events in the \kepler\ light curves is $\sim 60$.

This results is much more optimistic than the pre-launch estimate from
\citet{Yee:2008} for a few reasons.
One effect is that, the \kepler\ Mission ran for more than four
years---longer than the fiducial Mission goal---and $\sim190,000$ stars were
targeted for nearly the full baseline instead of the original 100,000.
\dfmtodo{Explain the other reasons.}

@@

The Random Forest (RF) classification model \citep{Breiman:2001} is a popular
model in the machine learning community where it is a common go-to model for
basic classification tasks.
It has also been applied with great success in astronomy \citep[for
example][]{Richards:2011, Richards:2012, Jenkins:2014}.
The RF model works by fitting an ensemble of decision trees to randomly
selected subsamples of the training dataset.
Each tree in the forest is ``grown'' by greedily choosing decision boundaries
that optimize the separation of the training data in randomly selected
subsets of the features until the separation is complete, each leaf contains
only one class or, at most, a fixed number of samples.
At each branch, the decision boundary is set by maximizing either the entropy
or the ``gini'' coefficient.

@@

To generate the \texttt{transit} examples, we simulate the transit signal
induced by the physical orbit of a large planet.
The specific distribution of simulation parameters is listed in \tab{params}.
In these simulations, we take limb darkening and integration time into
account\footnote{\url{https://github.com/dfm/transit}}
\citep{Mandel:2002, Kipping:2010}.
For the purposes of this \paper, we only simulate circular orbits under the
assumption that a single transit of a planet on an elliptical orbit would be
nearly identical to a circular orbit at a different period.

@@

\begin{table}[htbp]
\begin{center}
\begin{tabular}{lcc}
\toprule
Parameter & Units & Distribution \\
\midrule

limb darkening parameters $q_1$ and $q_2$ & --- & $q \sim U(0,\,1)$ \\
orbital period \period & days & $\ln \period \sim U(\ln 1500,\,\ln 5000)$ \\
reference transit time $\delta t$ & days & $\delta t \sim U(-0.15,\,0.15)$ \\
planet radius \rp & $R_\oplus$ & $\ln \rp \sim U(\ln 5,\,\ln 20)$ \\
impact parameter $b$ & --- & $b \sim U(0,\,1)$ \\

\bottomrule
\end{tabular}
\end{center}
\caption{%
The distribution of physical parameters for the injected signals.
The limb darkening parameterization is given by \citet{Kipping:2013}.
\tablabel{params}}
\end{table}

@@

% \dfmtodo{make a figure}

@@

In many supervised classification problems, a sophisticated feature
extraction method is used to improve the performance of the procedure.
For this specific problem, we find excellent performance using the 201
normalized flux values as the features.
The depth of the transit and the noise in the light curve is not irrelevant
to detection so we don't normalize the features to unit variance in the
standard way.
Instead, we normalize each sample (feature vector) by it's empirical median
value and take the logarithm.
Another option for a feature set could be the Fourier transform or wavelet
transform of the light curve section.
Alternatively, the spectrogram of the fluxes in a larger window.
We have experimented briefly with these options but find similar performance
on the cases that we tested.

@@

As discussed previously, recall is the probability, integrated across the full
parameter space of interest that a true transit signal will be detected by the
method.
In astronomy, this quantity is generally called the completeness or detection
efficiency and it is routinely measured for transit surveys
\citep{Petigura:2013, Dressing:2015, Foreman-Mackey:2015}.
On the other hand, precision is the probability that a positive classification
will be true.
In astronomy, precision is generally hard to measure because we rarely have
access to the ground truth.
For this problem, correctly computing the recall is impossible because we
never know that \emph{there is definitely no transit} at a given time in the
light curve.

@@

{\item \emph{the range of physical parameters used to simulate the transit
signals} --- The search will be most sensitive to transits with parameters
spanned by the training set but training on too many low signal-to-noise
examples in noisy light curves leads to decreased precision.
The parameter ranges used for the demonstrations here are given in
\tab{params}.}

@@

We estimate the precision of each RF classification model on the validation
set as described in \sect{train}.
As discussed in that \sectionname, we choose a target precision to
100~percent and set the score threshold accordingly.
This estimate is only applicable to the test set under two major assumptions.
The first is that all transiting planets with more than one transit and a
large signal-to-noise ratio have been previously discovered by one of the many
transit searches applied the \kepler\ dataset \citep{Burke:2014}.
This is a safe assumption because, for large planets, these surveys are all
largely complete out to periods yielding only 2 or 3 transits in the \kepler\
baseline \citep[for example][]{Petigura:2013}.
The second assumption is not always valid so it makes this estimate of the
precision only approximate and not completely reliable.
This assumption is that the noise processes in the data are stationary.
In other words, the transit-like noise signals in the training and validation
sets must be ``similar'' to the signals in the test section.
This constraint is nearly satisfied when the PDC light curves are used
(\sect{data}) and the data are split carefully (\sect{split}).
As demonstrated by the mis-classifications shown in \sect{demo}, however, it
is clear that violations to this rule do exist in the data and final vetting
of the candidates must be applied using a physical understanding of the
signals.

@@

\dfmtodo{Add some discussion of the failures and make suggestions for better
vetting techniques.}

\begin{figure}[p]
\begin{center}
\includegraphics[width=0.5\textwidth]{figures/candidates.pdf}
\end{center}
\caption{%
Some representative light curve sections that were labeled as candidates by
the single transit search.
The light curve in the top-left panel is a convincing transit signal but the
other panels all appear to be caused by stellar variability or instrumental
effects.
\figlabel{candidates}}
\end{figure}

@@

In the 3500 light curves we searched for this \paper, we discovered one
convincing single transit signal at $830.8093\pm0.0002$~KBJD in the light
curve of the 14.9~Kep-mag G-dwarf KIC 10602068.
Despite the fact that this is a very large signal, its discovery has never
been reported.
Given the estimate of $\sim 60$ detectable single transits in the full
\kepler\ archival dataset, we do expect about one discovery in 3500 light
curves.

Even though only one transit is observed, if we assume that the transit is
caused by a body on a Keplerian orbit around the host star, we can place
constraints on the physical properties---even the orbital period---of the
transiting candidate.
This technique is similar in spirit to the ``photoeccentric effect''
\citep{Dawson:2012} and ``asterodensity profiling'' \citep{Kipping:2012}.
The fundamental principle is that the transit duration is a measurement of
the instantaneous velocity of the planetary motion and this yields a
constraint on the orbital period when combining this with a measurement of the
stellar density and a prior on the eccentricity of the orbit.

The photometrically derived physical properties for this star place it as a
G-dwarf with a mass and radius of about 90~percent Solar \citep{Huber:2014}.
Using these constraints and a beta function prior on the eccentricity of the
orbit \citep{Kipping:2013a}, we run Markov Chain Monte Carlo
\citep{Foreman-Mackey:2013} to sample the posterior probability for the
physical parameters of the system.
In this analysis, we take limb darkening and the finite exposure time into
account \citep{Mandel:2002, Kipping:2010, Kipping:2013}.
The results of this chain marginalized into the relevant physical dimensions
are shown in \fig{corner}.
The radius of the transiting body is measured to be $2.1 \pm
0.4\,R_\mathrm{J}$.
The period is very weakly constrained above a lower limit of
$760\,\mathrm{days}$---set by the fact that a second transit is not
detected---with 68~percent of the posterior mass in the range $760\,\mathrm{d}
< \period < 1347\,\mathrm{d}$.

Given the large radius of this candidate, it is probably stellar in nature
instead of planetary but this could be easily confirmed using radial velocity
follow-up.

@@

The discovery and characterization of transiting planets based on a single
transit event is crucial for the future of transiting exoplanet surveys.
Many of the most dynamically influential planets---like Jupiter in our Solar
system---exhibit only a single transit in the full observational baseline of
the \kepler.
This will become even more of a problem as upcoming surveys move to shorter
contiguous observations.
For example, the \tess\ Mission is planned to get full-sky coverage at
half-hour cadence but most of the sky will only be targeted for a month.
This means that even habitable zone planets orbiting cool stars will transit
their host \emph{at most once in the entire lifetime of the Mission}!

To date, no methods exist for systematically and robustly discovering single
transit events based on large photometric surveys.
In this \paper, we present a novel and conceptually unique solution to this
problem drawing on machine learning methods for supervised classification.
This method has immense potential because it can be designed to be very robust
to false positives and it can exploit the detailed shape of physical transits.

Despite the fact that single transits are unlikely even if these long-period
planets are intrinsically common, we estimate that $\sim 60$ events should be
detectable in the \kepler\ archival dataset by extrapolating recent models of
planet occurrence rates and taking selection effects into account.
When applied to 3500 light curves from the \kepler\ dataset, this method
recovers one previously unknown single transit event at
$830.8093\pm0.0002$~KBJD in the light curve of KIC 10602068.
This rate is consistent with the predicted yield of 60 events in the full
dataset of 190,000 light curves.

Assuming a bound Keplerian orbit, we place constraints on the physical
properties of this transit candidate KIC 10602068.01.
Using photometrically derived stellar properties, we find that this candidate
has a radius of $2.1 \pm 0.4\,R_\mathrm{J}$ placing it as a very large planet
or brown dwarf or a small star.

This method for transit search is built using supervised classification and
its performance relies on several strong assumptions about the datasets and
these assumptions are sometimes violated leading to some transit-like signals
that appear to be caused by noise to be misclassified as transits.
The most severe assumption is that the noise properties of the data are
stationary.
In other words, we assume that variability in one subset of a light curve is
completely spanned by the variability in the other sections.
This assumption is, in general, false because the stochastic processes that
cause stellar variability are complicated and non-stationary and the detector
is plagued by non-negligible catastrophic changes in sensitivity and response.
We attempt to mitigate this problem by using light curves that have been
preprocessed to remove most of the instrumental effects and carefully dividing
the data into subsets but some false positives are still incorrectly
identified as candidates.

One possible method for reducing the false positive rate would be to augment
the training dataset using heuristic simulations of common false positives or
the light curves of ``similar'' stars.
Another option is to recognize that this search drastically reduces the
parameter space requiring evaluation and comparing the predictive power of
a transit model to other heuristic models including stellar variability or
instrumental effects.
